Space-time structure may be topological and not geometrical

### 2020-02-21 Proceedings/2019-ICNFP/Carcassi2019ICNFPProceedings.tex
Update Carcassi2019ICNFPProceedings.tex

This idea also clarifies the division between topology and geometry. Topological constructs are more primitive than geometrical ones because they have no notion of size. Physically, it means that before we can assign distances between two points, we must be able to distinguish them, to tell them apart. The topology, then, comes first, as it tells us if and how the points are distinguished experimentally. This insight also tells us why different spaces, though topologically equivalent, have different geometry. In everyday space, the distance along any direction can be measured in the same unit, say meters. The geometry is Riemmanian. On phase space, within a degree of freedom, units of position cannot be compared to unit of momentum, therefore we cannot define angles. Yet, phase-space areas for a degree of freedom are proportional to the number of possible configurations. The fact that we can quantify areas for each degree of freedom is what gives us symplectic geometry. If we imagine for example the space of all possible blood work results, however, we have neither notion, and therefore we simply have a manifold that is not geometric.

For those interested in the ultimate structure of space-time these insights naturally lead to the following line of inquiry. Before we want to understand whether space-time is geometrically truly a Riemmanian manifold, we need to understand whether space-time is topologically truly a manifold. A manifold, simply put, is a set of possible cases $X$ that can be experimentally identified by a set of real values[TODO:ck that "real values" is what we want]. So, the question is: when can a set of possible cases, or possibilities for short, be experimentally identified by a real value? To be clear, the question is not when the results of our measurements are real numbers. The answer to this is simple: never. Our measurements are always finite precision. The question is when the cases we want to be able to distinguish are parameterized by the real line.

@@

% We can cite others from wikipedi
%F. Faltin, N. Metropolis, B. Ross and G.-C. Rota. The real numbers as a wreath product Advances in Mathematics, 16(1975), 278–304.
Mathematics offers many constructions of the real numbers from more primitive notions. Common techniques include completion through Cauchy sequences, using Dedekind cuts or adding constraints over an algebraic fields[TODO:ck singular or plural]]. In fact, Faltin et al. (TODO:cite) states:
\begin{quote}
Few mathematical structures have undergone as many revisions or have been presented in as many guises as the real numbers. Every generation re-examines the reals in the light of its values and mathematical objectives.
\end{quote}
Our present purpose is one such re-examination in the light of how these quantities are defined through experiments.

@@

A single reference is something that allows us to distinguish a before and an after, a smaller and a greater, a higher and a lower. We can formalize a reference with our more basic notion of a verifiable statement: a reference gives us two verifiable statements, ``the objects is before this reference'' and ``the object is after this reference''. This allows us to recast ordering relationship in terms of logical relationship. For example, if reference A is after reference B, then if ``the object is before reference B'' is true, ``the object is before reference A'' must also be true. And these logical relationships remain conceptually the same regardless of whether ``before'' means ``before in time'' or ``before along a spatial direction'', etc. These are the sort of basic ideas and language we use within our framework, and from these we work out a set of necessary and sufficient conditions that a set of references needs to satisfy so that they define a set of possible distinguishable cases, or possibilities for short, that is ordered like the real line.

The biggest obstacle in this undertaking, and possibly in fully appreciating it, is that our standard intuition traps us in circular arguments. For example, one issue is that references have an extent: they occupy some space and so do objects. If we are to measure the position of an object, then, it may extend before, on and after the mark of a ruler. In these cases we typically note the position of the beginning of the object and an [TODO: an or the?] end of the object. This implicitly relies on the fact that the resolution of our eyes is higher than that of the ruler, and that we can independently recognize parts of the object, namely the beginning and end. Identifying parts using a higher resolution corresponds, in the end, to new references. So, assuming this can always happen means implicitly assuming what we are tasked to derive.

Unfortunately, thinking visually and drawing pictures is inherently misleading because everything we draw is inherently ordered[TODO: use inherently twice?]. It confuses more than it helps, so much so that we ourselves progressed only when we stopped trying to reason with pictures and concentrated on just the boolean logic of our statements in truth tables. In that setting, the only elements we are allowed to define are those simple logical relationships we described before: if the object is before A, then it is not after B and it is before C. All properties and relationships between references, then, must be defined in that manner.

Using that strategy, we reached two major findings. The first major finding is that \emph{most of the conditions are not required by the real numbers specifically, but they are required by ordered quantities in general}. That is, the set of reference has to satisfy a few conditions just to be sure that the cases they distinguish experimentally are ordered (i.e.~can be thought to be all being one after the other). The conditions are the following.
\begin{itemize}
	\item The references must be strict [TODO: Isn't this after they've already been sufficiently refined?], meaning an object is either exclusively before, on or after the reference. No overlaps are allowed. In practice, it means that the extent of the object being measured is always much smaller than the extent of the reference, therefore it will always be considered wholly before, on or after the reference.
	\item The references must be aligned, meaning their before and after statements identify incremental regions. In practice, this means we can always distinguish between them and they remain with constant before/after relationships (e.g.~they do not fluctuate). 
	\item The references must be refinable, meaning we can always resolve overlaps and fill in the whole space. For example, if an object can be between two references, then we must be able to put a new reference between those two references as well. If we have two references that overlap, then we must be able to find other references that fit within that overlap, to tell what is before and after.
\end{itemize}
These are the conditions one must require to have any linear order, either discrete (which can be labeled by integers), continuous (which can be labeled by real numbers) or any other. The divergence between real and integer quantities lies in how many references one can put in between two references. In the integer case, we can only fit finitely many references. In the real case, we can fit infinitely many.

@@

The first thing we need to establish is the link between topology and experimental verifiability. Point-set topology (or general topology) is, knowingly or unknowingly, widely used in physics as it is the foundation of many other tools, such as differential geometry and Lie groups. Most mathematical structures used in physics are topological spaces. Therefore it is crucial to understand what physical content they capture.\footnote{In computer science, the link between topology and computability is already accepted [TODO: add ref?] though not widely known. There is a link between those concepts and the ones presented here, but, for brevity, we are not going to expand on it. The gist is that any computation device is also a physical system, and the output of a computation can be experimentally verified (i.e. we can read it).}

@@

We are not interested here to expand on these arguments or pick ones that are universally accepted. That will be the scope of future work. Here we simply want to show that it is relatively easy to formulate reasonable arguments. As long as one finds reasonable that one of (i-iv) fails, then our point follows: space-time structure at a fine scale will not be geometrical, will not be ordered, but it will still be topological, giving us some way to measure what can physically be well defined.[TODO: Check that this last sentence means what we want it to.  Maybe clarify what types of properties can still be physically well defined and distinguished/measured at these scales. Avoid the word measure, as it's usually associated with quantities?  Would a better word be distinguish?]

@@

We have seen what are the requirements to give rise to the real numbers through a set of of experimentally verifiable statements, which captures the most basic elements of how continuous quantities are measured in practice. We have also seen that those can only be considered idealized conditions, and that, when those requirements fail, ordering itself fails and no geometrical structure can be constructured. Naturally, this opens the question of what topological spaces would be appropriate in those regimes. This is something we do not have an answer for, though we can mentioned two of the possibly many scenarios.

@@

There is, however, a much more drastic scenario. General relativity tells us that the geometry depends on the content of the space (i.e. the energy-momentum tensor). What is in the space affects the notion of distance and angle. It would not be far fetched to assume that what is in the space also affects what can be distinguished experimentally. As the energy-matter distribution changes the geometry by curving space-time, it would, at a more fundamental level, change the topology as well. It is not clear to us how one would even start exploring such scenario.

@@

>>>>

This idea also clarifies the division between topology and geometry. Topological constructs are more primitive than geometrical ones because they have no notion of size. Physically, it means that before we can assign distances between two points, we must be able to distinguish them, to tell them apart. The topology, then, comes first, as it tells us if and how the points are distinguished experimentally. This insight also tells us why different spaces, though topologically equivalent, have different geometry. In everyday space, the distance along any direction can be measured in the same unit, say meters. The geometry is Riemmanian. On phase space, within a degree of freedom, units of position cannot be compared to units of momentum, therefore we cannot define angles. Yet, phase-space areas for a degree of freedom are proportional to the number of possible configurations. The fact that we can quantify areas for each degree of freedom is what gives us symplectic geometry. If we imagine for example the space of all possible blood work results, however, we have neither notion, and therefore we simply have a manifold that is not geometric.

For those interested in the ultimate structure of space-time these insights naturally lead to the following line of inquiry. Before we want to understand whether space-time is geometrically truly a Riemmanian manifold, we need to understand whether space-time is topologically truly a manifold. A manifold, simply put, is a set of possible cases $X$ that can be experimentally identified by a set of real values. So, the question is: when can a set of possible cases, or possibilities for short, be experimentally identified by a real value? To be clear, the question is not when the results of our measurements are real numbers. The answer to this is simple: never. Our measurements are always finite precision. The question is when the cases we want to be able to distinguish are parameterized by the real line.

@@

% We can cite others from wikipedi
%F. Faltin, N. Metropolis, B. Ross and G.-C. Rota. The real numbers as a wreath product Advances in Mathematics, 16(1975), 278–304.
Mathematics offers many constructions of the real numbers from more primitive notions. Common techniques include completion through Cauchy sequences, using Dedekind cuts or adding constraints over an algebraic field. In fact, Faltin et al. (TODO:cite) states:
\begin{quote}
Few mathematical structures have undergone as many revisions or have been presented in as many guises as the real numbers. Every generation re-examines the reals in the light of its values and mathematical objectives.
\end{quote}
Our present purpose is one such re-examination in the light of how these quantities are defined through experiments.

@@

A single reference is something that allows us to distinguish a before and an after, a smaller and a greater, a higher and a lower. We can formalize a reference with our more basic notion of a verifiable statement: a reference gives us two verifiable statements, ``the object is before this reference'' and ``the object is after this reference''. This allows us to recast ordering relationship in terms of logical relationship. For example, if reference A is after reference B, then if ``the object is before reference B'' is true, ``the object is before reference A'' must also be true. And these logical relationships remain conceptually the same regardless of whether ``before'' means ``before in time'' or ``before along a spatial direction'', etc. These are the sort of basic ideas and language we use within our framework, and from these we work out a set of necessary and sufficient conditions that a set of references needs to satisfy so that they define a set of possible distinguishable cases, or possibilities for short, that is ordered like the real line.

The biggest obstacle in this undertaking, and possibly in fully appreciating it, is that our standard intuition traps us in circular arguments. For example, one issue is that references have an extent: they occupy some space and so do objects. If we are to measure the position of an object, then, it may extend before, on and after the mark of a ruler. In these cases we typically note the position of the beginning of the object and of the end of the object. This implicitly relies on the fact that the resolution of our eyes is higher than that of the ruler, and that we can independently recognize parts of the object, namely the beginning and end. Identifying parts using a higher resolution corresponds, in the end, to new references. So, assuming this can always happen means implicitly assuming what we are tasked to derive.

Unfortunately, thinking visually and drawing pictures is misleading because everything we draw is inherently ordered. It confuses more than it helps, so much so that we ourselves progressed only when we stopped trying to reason with pictures and concentrated on just the boolean logic of our statements in truth tables. In that setting, the only elements we are allowed to define are those simple logical relationships we described before: if the object is before A, then it is not after B and it is before C. All properties and relationships between references, then, must be defined in that manner.

Using that strategy, we reached two major findings. The first major finding is that \emph{most of the conditions are not required by the real numbers specifically, but they are required by ordered quantities in general}. That is, the set of reference has to satisfy a few conditions just to be sure that the cases they distinguish experimentally are ordered (i.e.~can be thought to be all being one after the other). The conditions are the following.
\begin{itemize}
	\item The references must be strict, meaning that the cases before, on and after are mutually exclusive. In practice, if objects have an extent, this must be much smaller than the extent of the reference, such that it will always be considered wholly before, on or after the reference.
	\item The references must be aligned, meaning their before and after statements identify incremental regions. In practice, this means we can always distinguish between them and they remain with constant before/after relationships (e.g.~they do not fluctuate). 
	\item The references must be refinable, meaning we can always resolve overlaps and fill in the whole space. For example, if an object can be between two references, then we must be able to put a new reference between those two references as well. If we have two references that overlap, then we must be able to find other references that fit within that overlap, to tell what is before and after.
\end{itemize}
These are the conditions one must require to have any linear order, either discrete (which can be labeled by integers), continuous (which can be labeled by real numbers) or any other. The divergence between real and integer quantities lies in how many references one can put in between two references. In the integer case, we can only fit finitely many references. In the real case, we can fit infinitely many.

@@

The first thing we need to establish is the link between topology and experimental verifiability. Point-set topology (or general topology) is, knowingly or unknowingly, widely used in physics as it is the foundation of many other tools, such as differential geometry and Lie groups. Most mathematical structures used in physics are topological spaces. Therefore it is crucial to understand what physical content they capture.\footnote{In computer science, the link between topology and computability is already accepted [TODO: add ref?] though not widely known. There is a link between those concepts and the ones presented here, but, for brevity, we are not going to expand on it. The gist is that any computation device is also a physical system, and the output of a computation can be experimentally verified (i.e.~we can read it).}

@@

We are not interested here to expand on these arguments or pick ones that are universally accepted. That will be the scope of future work. Here we simply want to show that it is relatively easy to formulate reasonable arguments. As long as one finds reasonable that one of (i-iv) fails, then our point follows: space-time structure at a fine scale will not be geometrical, will not be ordered, but it will still be topological, giving us some way to experimentally distinguish what can physically be well defined.

@@

We have seen what the requirements are to give rise to the real numbers through a set of experimentally verifiable statements, which captures the most basic elements of how continuous quantities are measured in practice. We have also seen that those can only be considered idealized conditions, and that, when those requirements fail, ordering itself fails and no geometrical structure can be constructed. Naturally, this opens the question of what topological spaces would be appropriate in those regimes. This is something we do not have an answer for, though we can mention two of the possibly many scenarios.

@@

There is, however, a much more drastic scenario. General relativity tells us that the geometry depends on the content of the space (i.e.~the energy-momentum tensor). What is in the space affects the notion of distance and angle. It would not be far-fetched to assume that what is in the space also affects what can be distinguished experimentally. As the energy-matter distribution changes the geometry by curving space-time, it would, at a more fundamental level, change the topology as well. It is not clear to us at this time how one would even start exploring such a scenario.

@@

Funding for this work was provided in part by the MCubed program of the University of Michigan.
